An indexed construction of semi-simplicial and semi-cubical types

### 2022-06-08 paper/paper.tex
paper: uniform appendmask without bugs
xkeymask: introduce ifinmask to query mask; show implicits in { ... }
paper: major polish

@@

\title{An indexed construction of semi-simplicial and semi-cubical types}
\author{
  \textcolor{gray80}{Ramkumar Ramachandra} \\
  \itshape \textcolor{gray80}{Université Paris-Cité}
  \and
  \textcolor{gray80}{Hugo Herbelin} \\
  \itshape \textcolor{gray80}{Instit de Recherche en Informatique Fondamentale}
}
\date{}

@@

\section{Overture}
Cubical type theory~\cite{Bezem14}~\cite{Cohen16}~\cite{Angiuli17} is an extension of dependent type theory that provides a computational interpretation of Voevodsky's \emph{univalence axiom}, in a field known as \emph{homotopy type theory}, or HoTT. Two real-world projects that implement cubical type theory are worthy of mention: an experimental branch of Agda, known as \emph{Cubical Agda}~\cite{Vezzosi21}, and the Arend theorem prover~\footnote{\href{https://arend-lang.github.io}{arend-lang.github.io}}. Cubical type theory is usually modeled as cubical sets, in the set-theoretic setting. Efforts to model cubical type theory have been ongoing since 2014, and many of these efforts are in the set-theoretic setting.

Our contribution is the following. We have modelled the core of semi-cubical sets in a way that naturally generalizes to semi-simplicial sets in type theory, and we call our generalization \emph{$\nu$-sets}. In contrast to previous efforts that use a fibered representation, we use an indexed representation. Our work is, in principle axiom-free, although the accompanying artifact uses functional extensionality, and this requires disappears if $\nu$ is finite.

@@

The outline of the paper is as follows. First, we briefly describe semi-simplicial and semi-cubical sets, as it appears in mathematical literature. We then proceed to explain our construction, with the help of diagrams. Before spelling out the details of our formalization in type-theoretic language, we have a section on reasoning in sets, in which we spell out \texttt{HSet}.

\section{\texorpdfstring{$\nu$}{nu}-sets}
Simplicial sets are powerful mathematical objects, that form the basis of much of modern homotopy theory. Their power derives from the fact that their homotopy category is exceptionally well-behaved, and that there are multiple model structures defined on them. Unlike CW-complexes that rely on pure topological notions like spheres and disks to ``tame'' a topological space, the definition of simplicial sets is purely categorical, and exist independent of topological spaces. As such, simplicial sets exist purely in the imagination of the mathematician, as every simplicial set has an infinite number of degenerate simplices. What we have formalized is \emph{augmented} semi-simplicial sets, and semi-cubical sets, which is to say that they are the usual simplicial sets and cubical sets, minus the degeneracies.

@@

\section{Our construction}
The fibered representation can be drawn as:
\begin{equation*}
  \begin{tikzcd}
    X_0 : \U{} & X_1 : \U{} \arrow[l, "\delta^\epsilon_0" description, shift left=2] \arrow[l, "\delta^\omega_0" description, shift right=2] & X_2 : \U{} \arrow[l, "\delta^\epsilon_1" description, shift left=6] \arrow[l, "\delta^\epsilon_0" description, shift left=2] \arrow[l, "\delta^\omega_0" description, shift right=2] \arrow[l, "\delta^\omega_1" description, shift right=6] & \ldots
  \end{tikzcd}
\end{equation*}

The indexed representation can be drawn as:
\begin{align*}
  X_0 & : \U{}                                                                                                      \\
  X_1 & : X_0 \times X_0 \rightarrow \U{}                                                                           \\
  X_2 & : \forall a b c d, X_1 : ab \rightarrow X_1 : bc \rightarrow X_1 : cd \rightarrow X_1 : da \rightarrow \U{} \\
  \ldots
\end{align*}

\section{Reasoning about sets in type theory}
We dedicate this section to discussing UIP in different settings, and supply intuitions into this notion.

UIP is a flavor of proof-irrelevance:

@@

Proof irrelevance is an inherent part of set theory and first-order logic, although the property cannot be stated in the language of natural deductions. Intuititively, what it means is that a tree of deductions converges to a result, with no memory of the proof route; the tree itself is forgotten once the final result is obtained.

\section{Type-theoretic construction of \texorpdfstring{$\nu$}{nu}-sets}
In this section, we give the core of the definition of truncated $\nu$-sets
as a coinductive limit.

@@

The definition is dispatched over tables \ref{tab:core}, \ref{tab:frames}, \ref{tab:faces} and \ref{tab:coh}. It describes the structure of the underlying higher-dimensional relations on which fillers are built. Note the presence of a coherence condition $\cohf$ to ensuring that both sides of the equality in $\restrl$ and $\restrc$ are in the same type. The proof of $\cohf$ itself requires an higher-dimensional coherence condition which we obtain by working in HSet. Note that if the proofs of the same equality were not equated, there would be a need for arbitrary many higher-dimensional coherences (see e.g.~\cite{Herbelin15} for a discussion on the de facto need for recursive higher-dimensional coherence conditions in formulating higher-dimensional structures in type theory). Note also that for a given $n$, the coherence conditions evaluate to a reflexivity proof, so that the construction evaluates to an effective sequence of types of iterated relations not mentioning $\restrf$ nor $\cohf$ anymore.

We call the structure thus defined \emph{bare truncated sets}: \emph{bare} because it can be seen as defining truncated $\nu$-sets corresponding to simplical sets with only faces as part of the structure.

@@

% For just the second table, keep the universe letter in fullframe
\newcommandx{\fullframe}[3][1,2,3]{\prim{fullframe}[#1][#2][][][#3]}
\renewcommand*{\lab}{tab:core}
\input{tab-core.tex}

@@

% For the last two tables
\appendmask[bonak]{frame}[D]
\appendmask[bonak]{layer}[D]
\appendmask[bonak]{filler}[D, E]
\appendmask[bonak]{restrframe}[D]
\appendmask[bonak]{restrlayer}[D, d]
\appendmask[bonak]{restrfiller}[D, d]

@@

% The third table mentions cohframe; keep ε, ω in this instance
\renewcommandx{\cohf}[9][1,2,3,4,5,6,7,8,9]{\coh{frame}[][#2][#3][][][#6,#7][#8][#9]}
\renewcommand*{\lab}{tab:faces}
\input{tab-faces.tex}

% For the last table,
% (1) make the {d} arguments of restr implicit
% (2) make the {D, E} arguments of coh implicit
\appendmask[bonak]{cohframe}[D]
\appendmask[bonak]{cohlayer}[D]
\appendmask[bonak]{cohfiller}[D, E]

@@

% Coherence conditions for frame, layer, and filler
% Drop universe letter, ε, and ω
\renewcommandx{\cohf}[9][1,2,3,4,5,6,7,8,9]{\coh{frame}[][][][#4][#5][#6,#7][#8][#9]}
\renewcommandx{\cohl}[9][1,2,3,4,5,6,7,8,9]{\coh{layer}[][][][#4][#5][#6,#7][#8][#9]}
\renewcommandx{\cohc}[9][1,2,3,4,5,6,7,8,9]{\coh{filler}[][][][#4][#5][#6,#7][#8][#9]}
\renewcommand*{\lab}{tab:coh}
\input{tab-coh.tex}

@@

\newpage
\appendix
\section{Tables in full detail}
% Clear the mask and label
\clearmask

>>>>

% \eqnarg is slightly more involved: output ( ... ) if not in mask,
% and { ... } if in mask
\newcommandx*{\eqnarg}[3]{\ifinmask[bonak]{#1}[#2]{\{#2:#3\}}{(#2:#3)}}

@@

\title{An indexed construction of semi-simplicial and semi-cubical types}
\author{
  \textcolor{gray80}{Ramkumar Ramachandra} \\
  \itshape \textcolor{gray80}{Université Paris-Cité}
  \and
  \textcolor{gray80}{Hugo Herbelin} \\
  \itshape \textcolor{gray80}{Institut de Recherche en Informatique Fondamentale}
}
\date{}

@@

\section{Overture}
Cubical type theory~\cite{Bezem14}~\cite{Cohen16}~\cite{Angiuli17} is an extension of dependent type theory that provides a computational interpretation of Voevodsky's \emph{univalence axiom}, in a field known as \emph{homotopy type theory}, or HoTT. The most significant implementation of cubical type theory in the real world is an experimental branch of Agda, known as \emph{Cubical Agda}~\cite{Vezzosi21}. Cubical type theory is usually modeled as cubical sets, in the set-theoretic setting.

Our contribution is the following. We have modelled the core of semi-cubical sets in a way that naturally generalizes to augmented semi-simplicial sets in type theory, and we call our generalization \emph{$\nu$-sets}. In contrast to previous efforts that use a fibered representation, we use an indexed representation. Our work is, in principle axiom-free, although the accompanying artifact uses functional extensionality, and this requires disappears if $\nu$ is finite.

@@

The outline of the paper is as follows. First, we briefly describe augmented semi-simplicial and semi-cubical sets, as it appears in mathematical literature, and our generalization to $\nu$-sets. We then proceed to explain our construction, with the help of diagrams. Before spelling out the details of our formalization in type-theoretic language, we have a section on reasoning in sets.

\section{Mathematical treatment of \texorpdfstring{$\nu$}{nu}-sets}
It is well-known that simplicial sets exist purely in the imagination of the mathematician, as every simplicial set has an infinite number of degenerate simplices. What we have formalized is \emph{augmented} semi-simplicial sets, and semi-cubical sets, which is to say that they are the usual simplicial sets and cubical sets, minus the degeneracies. Since they're not of particular interest, and indeed, we plan to extend our work to do degeneracies in the future, we briefly explain the mathematical objects we have formalized.

@@

\subsection{Generalization to \texorpdfstring{$\nu$}{nu}}

\begin{table}[H]
  \begin{tabularx}{\linewidth}{p{.2\linewidth}|p{.4\linewidth}|p{.3\linewidth}}
    \toprule
    Value of $\nu$ & 1                              & 2                  \\
    \graymidrule
    Interpretation & Augmented semi-simplical types & Semi-cubical types \\
    \bottomrule
  \end{tabularx}
\end{table}

\section{Our construction}
In this section, we briefly explain the difference between the fibered and indexed representations, and illustrate our approach to the construction, with the help of some figures.

The fibered representation can be illustrated as:
\begin{equation*}
  \begin{tikzcd}
    X_0 : \U{} & X_1 : \U{} \arrow[l, "\delta^\epsilon_0" description, shift left=2] \arrow[l, "\delta^\omega_0" description, shift right=2] & X_2 : \U{} \arrow[l, "\delta^\epsilon_1" description, shift left=6] \arrow[l, "\delta^\epsilon_0" description, shift left=2] \arrow[l, "\delta^\omega_0" description, shift right=2] \arrow[l, "\delta^\omega_1" description, shift right=6] & \ldots
  \end{tikzcd}
\end{equation*}

The indexed representation can be illustrated as:
\begin{align*}
  X_0 & : \U{}                                                                                                      \\
  X_1 & : X_0 \times X_0 \rightarrow \U{}                                                                           \\
  X_2 & : \forall a b c d, X_1 : ab \rightarrow X_1 : bc \rightarrow X_1 : cd \rightarrow X_1 : da \rightarrow \U{} \\
  \ldots
\end{align*}

One way to obtain one construction from the other is via the \emph{Grothendieck construction}, although this is not relvant to our work.

\subsection{Iterated parameteriticity}

\section{Reasoning about sets in type theory}
Since proof irrelevance is an inherent part of set theory and first-order logic, and proof irrelevance is \emph{extensional} in type theory, we work in a universe known as \texttt{HSet}, which we describe shortly.

\subsection{Unicity of identity proofs}
UIP is a flavor of proof-irrelevance:

@@

In type theory, UIP can be proved on the unit type:

\begin{align*}
  \mathsf{eq\_refl}        & : \forall x, x = x                               \\
  \mathsf{UIP\_refl\_unit} & : \forall x : \top = \top, x = \mathsf{eq\_refl}
\end{align*}

\subsection{HSet}
\texttt{HSet} provides us a restricted setting of UIP, and our implementation of \texttt{HSet} is a straightforward packaging of two fields:

\begin{align*}
  \mathsf{Dom} & : \mathsf{Type}                                          \\
  \mathsf{UIP} & : \forall x y : \mathsf{Dom}, \forall p q : x = y, p = q
\end{align*}

In the \texttt{HSet} universe, three properties hold:

\begin{enumerate}
  \item[(i)] UIP holds on the unit type.
  \item[(ii)] UIP propogates to $\Sigma$-types.
  \item[(iii)] UIP propogates to $\Pi$-types, with some additional functional extensionality axioms.
\end{enumerate}

\section{Type-theoretic construction of \texorpdfstring{$\nu$}{nu}-sets}
% Abbreviated tables in this section
%
% The rule for implicit arguments for each table:
% show 1 explicit in first two parts
% show E + 1 explicit in the third part
\appendmask[bonak]{layer}[D]
\appendmask[bonak]{filler}[D]
\appendmask[bonak]{restrframe}[D]
\appendmask[bonak]{restrlayer}[D, d]
\appendmask[bonak]{restrfiller}[D, d]
\appendmask[bonak]{cohframe}[D]
\appendmask[bonak]{cohlayer}[D, d]
\appendmask[bonak]{cohfiller}[D, d]

@@

The definition in \ref{tab:coind}, the coinductive limit, is dispatched over tables \ref{tab:core}, \ref{tab:frames}, \ref{tab:faces} and \ref{tab:coh}. They describe the structure of the underlying higher-dimensional relations on which fillers are built.

@@

% For just the second table, keep the universe letter in fullframe
\newcommandx{\fullframe}[3][1,2,3]{\prim{fullframe}[#1][#2][][][#3]}

We now describe \emph{bare truncated sets}: \emph{bare} because it can be seen as defining truncated $\nu$-sets without face maps, and \emph{truncated} because it is $n$-truncated.

\renewcommand*{\lab}{tab:core}
\input{tab-core.tex}

@@

@@

% Hack to remove , in \cohf
\renewcommandx{\cohf}[9][1,2,3,4,5,6,7,8,9]{\coh{frame}[][#2][#3][][][][][#9]}

In the way we have chosen to formlate these tables, there is, for instance, coherence condition $\cohf$ both sides of the equality in $\restrl$ and $\restrc$, to ensure that they are the same type.

For a given $n$, the coherence conditions evaluate to a reflexivity proof, so that the construction evaluates to an effective sequence of types of iterated relations not mentioning $\restrf$ nor $\cohf$ anymore.

% The third table mentions cohframe; keep ε, ω in this instance
\renewcommandx{\cohf}[9][1,2,3,4,5,6,7,8,9]{\coh{frame}[][#2][#3][][][#6,#7][#8][#9]}

\renewcommand*{\lab}{tab:faces}
\input{tab-faces.tex}

@@

% Hack to remove , in \cohf
\renewcommandx{\cohf}[9][1,2,3,4,5,6,7,8,9]{\coh{frame}[][#2][#3][][][][][#9]}

The proof of $\cohf$ require an higher-dimensional coherence condition which we obtain by working in \texttt{HSet}. If the proofs of the same equality were not equated, there would be a need for arbitrary many higher-dimensional coherences (see e.g.~\cite{Herbelin15} for a discussion on the de facto need for recursive higher-dimensional coherence conditions in formulating higher-dimensional structures in type theory).

% Coherence conditions for frame, layer, and filler
% Drop universe letter, ε, and ω
\renewcommandx{\cohf}[9][1,2,3,4,5,6,7,8,9]{\coh{frame}[][][][#4][#5][#6,#7][#8][#9]}
\renewcommandx{\cohl}[9][1,2,3,4,5,6,7,8,9]{\coh{layer}[][][][#4][#5][#6,#7][#8][#9]}
\renewcommandx{\cohc}[9][1,2,3,4,5,6,7,8,9]{\coh{filler}[][][][#4][#5][#6,#7][#8][#9]}

\renewcommand*{\lab}{tab:coh}
\input{tab-coh.tex}

@@

\newpage
\appendix
\section{Tables in full detail}
% Full tables in this section
%
% Clear the mask and label
\clearmask
